---
title: "Ridge and LASSO Modeling for DC Crime"
subtitle: "Kogod School of Business"
author: "Assem Mendygaziyeva"
date: "2026-02-16"
format:
  html: default
  docx: default
---

## 

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=T, warning=FALSE, message=FALSE)
```

```{r}
setwd("/Users/assemmendygaziyeva/Desktop")
crime <- read.csv("all_crime_incidents_2020-2026.csv")

head(crime)
```

# **Business Case and Analytic Question**

The project focuses on helping DC public safety agencies decide where and when to prioritize policing, environmental design, and social programs based on recent crime patterns. The team analytic question is: *How do temporal features (such as month and shift) and location characteristics (such as ward and neighborhood areas) jointly predict crime levels and types, and how can these patterns inform deployment recommendations?* In this Deliverable, ridge and LASSO regression models are used to predict monthly crime counts by ward and crime type, in order to identify which wards and crime categories consistently drive higher incident volumes and to support more targeted resource allocation decisions.

# **Data Preparation and Cleaning**

This analysis draws on the combined 2020–2026 DC incident‑level dataset, which contains one row per reported crime and fields such as REPORT_DAT, OFFENSE, WARD, SHIFT, and location identifiers. REPORT_DAT is converted to a Date variable (DATE), WARD and SHIFT are cast as factors, and rows with missing ward values are removed. Incidents are then summarized by month (using floor_date(DATE, "month")), WARD, SHIFT, and OFFENSE to obtain a TOTAL_INCIDENTS count for each group. OFFENSE is recoded into four broader crime_type groups (homicide, robbery_burglary, theft, and sex_abuse/arson/assault) and the data are aggregated again by month, crime_type, and WARD to produce the modeling table df_counts, where each row contains the total_crime_count for that month–crime_type–ward combination. In the ridge and LASSO models, month, crime_type, and WARD are used as predictors, so SHIFT is only used in the intermediate aggregation, not as a final predictor.

```{r}

library(tidyverse)
library(lubridate)

# Create DATE, factor WARD and SHIFT
all_crimes <- crime |>
  mutate(
    DATE = as.Date(REPORT_DAT),
    WARD = as.factor(WARD),
    SHIFT = as.factor(SHIFT)
  )

# Variables used for initial counts
group_vars <- c("WARD", "SHIFT", "OFFENSE")

# Count incidents by month, ward, shift, offense
df_counts_summarized <- all_crimes |>
  filter(!is.na(WARD)) |>
  group_by(
    month = floor_date(DATE, "month"),
    across(all_of(group_vars))
  ) |>
  summarise(
    TOTAL_INCIDENTS = n(),
    .groups = "drop"
  )

# Recode OFFENSE into four crime_type groups and sum to total_crime_count
df_counts <- df_counts_summarized |>
  mutate(
    crime_type = case_when(
      str_detect(OFFENSE, "THEFT F/AUTO|THEFT/OTHER|MOTOR VEHICLE THEFT") ~ "theft",
      str_detect(OFFENSE, "HOMICIDE") ~ "homicide",
      str_detect(OFFENSE, "ROBBERY|BURGLARY") ~ "robbery_burglary",
      TRUE ~ "sex_abuse/arson/assault"
    ),
    crime_type = factor(
      crime_type,
      levels = c("homicide", "robbery_burglary", "theft", "sex_abuse/arson/assault")
    )
  ) |>
  group_by(month, crime_type, WARD) |>
  summarise(
    total_crime_count = sum(TOTAL_INCIDENTS),
    .groups = "drop"
  )


```

```{r}
library(glmnet)

# Model matrix with dummy variables (glmnet needs numeric x)
x <- model.matrix(
  total_crime_count ~ month + crime_type + WARD,
  data = df_counts
)[, -1]  # drop intercept column

# Outcome vector
y <- df_counts$total_crime_count

# Train/test split (70% train, 30% test)
set.seed(621)
n <- nrow(df_counts)
train_idx <- sample(1:n, size = floor(0.7 * n))

x_train <- x[train_idx, ]
y_train <- y[train_idx]

x_test  <- x[-train_idx, ]
y_test  <- y[-train_idx]

```

# Baseline Model

```{r}
set.seed(621)
n <- nrow(df_counts)
train_idx <- sample(1:n, size = floor(0.7 * n))

train_df <- df_counts[train_idx, ]
test_df  <- df_counts[-train_idx, ]

ols_model <- lm(total_crime_count ~ ., data = train_df)

ols_pred <- predict(ols_model, newdata = test_df)
ols_rmse <- sqrt(mean((test_df$total_crime_count - ols_pred)^2))
ols_rmse
```

# Model 1: Ridge (lambda.min)

Mild shrinkage, keeps all variables

```{r}
# Cross‑validated ridge (alpha = 0)
cvridge <- cv.glmnet(
  x_train,
  y_train,
  alpha  = 0,          # ridge
  family = "gaussian"
)

#plot
plot(cvridge)

#Best lambda and its CV error
lambda_ridge_min <- cvridge$lambda.min
min_cv_ridge     <- min(cvridge$cvm)

cbind(
  best_lambda_ridge = lambda_ridge_min,
  best_cvm_ridge    = min_cv_ridge
)

#Fit ridge at lambda.min and show coefficients
ridge_min <- glmnet(
  x_train,
  y_train,
  alpha  = 0,
  lambda = lambda_ridge_min,
  family = "gaussian"
)
coef(ridge_min)

#Test RMSE
ridge_min_pred <- predict(ridge_min, newx = x_test)
ridge_min_rmse <- sqrt(mean((y_test - ridge_min_pred)^2))
ridge_min_rmse

```

# Model 2: Ridge (lambda.1se)

Stronger shrinkage, but still keeps all variables.

```{r}
#Lambda.1se from same CV object
lambda_ridge_1se <- cvridge$lambda.1se
lambda_ridge_1se

#Ridge model at lambda.1se and coefficients
ridge_1se <- glmnet(
  x_train,
  y_train,
  alpha  = 0,
  lambda = lambda_ridge_1se,
  family = "gaussian"
)
coef(ridge_1se)

#RMSE
ridge_1se_pred <- predict(ridge_1se, newx = x_test)
ridge_1se_rmse <- sqrt(mean((y_test - ridge_1se_pred)^2))
ridge_1se_rmse

```

# Model 3: LASSO (lambda.min)

Shrinkage + can set small effects close to zero

```{r}

# Cross‑validated LASSO (alpha = 1)
cvlasso <- cv.glmnet(
  x_train,
  y_train,
  alpha  = 1,          # LASSO
  family = "gaussian"
)

#plot
plot(cvlasso)

#Best lambda and its CV error
lambda_lasso_min <- cvlasso$lambda.min
min_cv_lasso     <- min(cvlasso$cvm)

cbind(
  best_lambda_lasso = lambda_lasso_min,
  best_cvm_lasso    = min_cv_lasso
)

#Fit LASSO at lambda.min and show coefficients
lasso_min <- glmnet(
  x_train,
  y_train,
  alpha  = 1,
  lambda = lambda_lasso_min,
  family = "gaussian"
)
coef(lasso_min)

#RMSE
lasso_min_pred <- predict(lasso_min, newx = x_test)
lasso_min_rmse <- sqrt(mean((y_test - lasso_min_pred)^2))
lasso_min_rmse

```

# Model 4: LASSO (lambda.1se)

Even stronger penalty, drops more variables (more zeros)

```{r}
#Lambda.1se from LASSO CV
lambda_lasso_1se <- cvlasso$lambda.1se
lambda_lasso_1se

#LASSO at lambda.1se and coefficients
lasso_1se <- glmnet(
  x_train,
  y_train,
  alpha  = 1,
  lambda = lambda_lasso_1se,
  family = "gaussian"
)
coef(lasso_1se)

#RMSE
lasso_1se_pred <- predict(lasso_1se, newx = x_test)
lasso_1se_rmse <- sqrt(mean((y_test - lasso_1se_pred)^2))
lasso_1se_rmse

```

In the LASSO (lambda.1se) model, the largest positive coefficient is for theft, confirming that theft-related incidents drive the highest monthly volumes across wards, while robbery/burglary also increases expected counts but to a smaller degree. Several ward effects remain non‑zero (e.g., Ward 2 positive, Wards 3 and 4 negative), indicating that some wards consistently experience higher or lower crime counts than the reference ward after controlling for crime type and time.

# Compare all four models

RMSE comparison across models

```{r}
rmse_results <- tibble(
  model = c(
    "Ridge (lambda.min)",
    "Ridge (lambda.1se)",
    "LASSO (lambda.min)",
    "LASSO (lambda.1se)",
    "Baseline Model"
  ),
  RMSE = c(
    ridge_min_rmse,
    ridge_1se_rmse,
    lasso_min_rmse,
    lasso_1se_rmse,
    ols_rmse
  )
)

rmse_results

```

# **Model Performance Summary**

Across the four regularized models, test RMSE values range from about 41.1 to 42.8, indicating broadly similar predictive performance. LASSO with lambda.min achieves the lowest RMSE (about 41.1), followed closely by ridge with lambda.min (about 41.6), while the stronger‑penalty models (ridge and LASSO with lambda.1se) have slightly higher RMSEs around 42–42.8. These results suggest that all four specifications capture the main crime patterns similarly well when using month, crime_type, and WARD as predictors. The baseline OLS model already achieves strong predictive performance (RMSE of about 41.1), and the ridge/LASSO models deliver very similar RMSE, indicating that regularization does not materially improve out‑of‑sample accuracy for the aggregated crime counts. However, the LASSO models—especially with the stronger lambda.1se penalty—shrink small coefficients toward zero and drop weaker ward/crime_type effects, providing a sparser, more interpretable set of drivers of crime volumes while maintaining comparable predictive power.

# **Recommended Model and Rationale**

Four regularized models (two ridge, two LASSO) are estimated to predict monthly crime counts by ward and crime type, all using the same predictors (month, crime_type, WARD). Their test RMSE values are very similar: they all miss by about 41–43 crimes on average, with LASSO lambda.min at about 41.1 and LASSO lambda.1se at about 42.3. Because these accuracy numbers are so close, it is reasonable to treat the models as having similar predictive performance and to prioritize simplicity. The LASSO (lambda.1se) model applies a stronger penalty, which shrinks more coefficients to zero and drops month and several crime_type/ward effects, keeping only the strongest signals—especially theft, robbery/burglary, and a subset of wards (2, 3, 4, 5, 8) with consistently higher or lower crime. This simpler model still predicts about as well as the more complex versions but gives a clearer story for decision‑makers, highlighting which crime types and wards matter most for planning deployments without being cluttered by small, noisy effects. For these reasons, LASSO with lambda.1se is recommended as the final model, as it balances good predictive accuracy with a more interpretable and operationally useful set of drivers of crime counts.
